---
title: "Exploring the Impact of Dietary and Environmental Factors on Hypertension and Mercury Exposure"
author: "Danish Maknojia, Mona Saeed, Leonard Eshun, Su Zhang"
subtitle: "2024-12-08"
format: pdf
header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
      - \usepackage{booktabs}
      - \usepackage{caption}

execute:
  echo: false
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
library(haven)  
library(tidyr)
library(hoopR)
library(foreign)   
library(dplyr)
library(tidyr)
library(MASS)
library(nnet)
library(car)
library(brant)
library(pscl)
library(tidyverse)
library(caret)
library(modelsummary)
library(knitr)
library(kableExtra)
library(skimr)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Abstract

Investigating factors influencing blood pressure and mercury levels in the US population is crucial for public health interventions and environmental policy-making. This study aims to examine the relationships between dietary sodium intake and blood pressure levels, as well as factors that influence blood mercury levels. The dataset analyzed is derived from the Center for Disease Control (CDC) survey data collected through interviews, blood tests, and medical examinations from August 2021 to August 2023. Multinomial regression was used to model blood pressure levels, and linear regression was applied to model blood mercury levels. The analysis revealed no significant association between dietary sodium consumption and blood pressure levels. However, age emerged as a significant predictor of hypertension (p \< 0.001). Older people are more likely to have hypertension. After preliminary data analysis and variable selection, tap water consumption showed a significant positive association with blood mercury levels (p \< 0.001). Race and income demonstrated significant interactive effects, with non-Hispanic Asian participants showing notably higher mercury levels (p \< 0.001). Our study identified key environmental and demographic factors affecting blood mercury levels, which can inform public health policies regarding water consumption and exposure to environmental toxins.

## Introduction

Hypertension and environmental exposure to mercury are significant public health concerns worldwide. With hypertension, affecting over 1.28 billion adults worldwide with a majority in low- and middle-income countries, and environmental mercury exposure, linked to cardiovascular and neurological issues such as elevated blood pressure even at low levels, are significant global public health concerns (World Health Organization; Houston, p 621). Understanding the factors that influence these conditions is critical to improving health outcomes and developing targeted interventions. This study leverages data from the U.S Centers for Disease Control and Prevention (CDC), a nationally representative dataset, to explore two key research questions that address these pressing health issues.

The first research question examines the association between dietary sodium intake and blood pressure levels, with a particular focus on whether the effect varies by factors such as gender and race. Blood pressure is an ordinal variable, categorized into four levels according to CDC guidelines, ranging from normal to Stage 2 hypertension (CDC, 2023). High dietary sodium intake has long been implicated as a risk factor for hypertension, a leading contributor to cardiovascular diseases globally (World Health Organization, 2021). Gender and race may further modulate this relationship due to physiological and social determinants of health, making it essential to explore these interactions. Answering this question provides insights that can inform dietary guidelines and public health policies aimed at reducing hypertension-related morbidity and mortality.

The second research question investigates the impact of drinking water sources, such as tap or bottled water, on blood mercury levels. Blood mercury is a continuous variable that reflects exposure to mercury, a toxic heavy metal with known adverse effects on neurological and cardiovascular health (Agency for Toxic Substances and Disease Registry, 2023). Factors such as race may influence this relationship, potentially due to disparities in environmental exposure or differences in water consumption patterns. This question is vital because understanding the relationship between water sources and mercury exposure can guide environmental regulations and public health recommendations, particularly in vulnerable populations.

These research questions address two critical dimensions of health: dietary and environmental factors. By exploring the interplay between dietary sodium, drinking water sources, and key demographic factors, this study aims to provide actionable insights into mitigating the risks of hypertension and mercury toxicity. The findings have the potential to inform individual behavioral changes, public health interventions, and broader environmental and dietary policies.

## Methods

### Data Overview and Preprocessing

The dataset used for the first research question consists of information on 7,801 observations, including variables such as sodium intake, blood pressure categories, gender, race, age, and alcohol consumption. The outcome variable, **blood pressure category**, is ordinal and categorized as *Normal*, *Elevated*, *Stage 1 Hypertension*, and *Stage 2 Hypertension* based on both the systolic and diastolic blood measurements as illustrated in the table below. There were 1830 observations with missing values for alcohol consumption and sodium intake and 323 observations have missing values in systolic and diastolic blood pressure measurements.

```{r, results='hide'}

# Load datasets
BP   <- read_xpt("https://github.com/siliconshells/mids_stats_group_ten/raw/refs/heads/main/00_source_data/cdc_data/BPXO_L.XPT")
demographics   <- read_xpt("https://github.com/siliconshells/mids_stats_group_ten/raw/refs/heads/main/00_source_data/cdc_data/DEMO_L.XPT")
mercury   <- read_xpt("https://github.com/siliconshells/mids_stats_group_ten/raw/refs/heads/main/00_source_data/cdc_data/PBCD_L.XPT")
dietary <- read_xpt("https://github.com/siliconshells/mids_stats_group_ten/raw/refs/heads/main/00_source_data/cdc_data/DR1TOT_L.XPT")

# Merge datasets on sequence number (SEQN)
data_merged_q1 <- demographics %>%
  inner_join(BP, by = "SEQN") %>%
  inner_join(dietary, by = "SEQN") %>%
  inner_join(mercury, by = "SEQN")

# Pick only the columns of interest
data_q1_3_measures <- data_merged_q1 %>%
  dplyr::select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH1, RIDRETH3, DR1TSODI, BPXOSY1, BPXOSY2, BPXOSY3, BPXODI1, BPXODI2, BPXODI3, DR1TALCO)

print(colSums(is.na(data_q1_3_measures)))
```

| Blood Pressure Category | Systolic Blood Pressure | Diastolic Blood Pressure |
|------------------------|------------------------|-------------------------|
| Normal                  | \<120 mmHg              | and \<80 mmHg            |
| Elevated                | 120-129 mmHg            | and \<80 mmHg            |
| Stage 1 Hypertension    | 130-139 mmHg            | or 80-89 mmHg            |
| Stage 2 Hypertension    | ≥140 mmHg               | or ≥90 mmHg              |

The systolic and diastolic blood measurements used in this analysis represent the mean of the three measurements taken for each participant in the original dataset. These measurements were taken on three consecutive days. Predictor variables include continuous variables (e.g., sodium intake, age, alcohol consumption) and categorical variables (e.g., gender, race).

For the second research question on blood mercury level, the data has 7785 observations, among which 2870 observations have missing values in either predictor or outcome variables. This dataset has missing values in variables of mercury level, income level, and water intakes of bottled water and tap water.

Stochastic regression imputations were used to handle missing values because (1) dataset size is large enough to build reliable imputation models; (2) missing values appear in both predictors and outcome; (3) maintains the natural associations in the data; (4) better than median or mean imputations, which are more arbitrary and underestimate variance. Extreme outliers in sodium intake and alcohol consumption were identified using visualizations (boxplots) and summary statistics. Sensitivity analyses were planned to evaluate the impact of including these points. Descriptive statistics (mean, median, range) and visualizations (histograms and boxplots) were used to summarize continuous variables. Bar charts and frequency tables were used to explore the distribution of categorical variables. Relationships between sodium intake and blood pressure levels were explored using boxplots and correlation analyses. Associations between other predictors (e.g., gender, age, race) and blood pressure categories were examined.

```{r, results='hide'}
# take the mean of the three blood pressure measurements (both systolic and diastolic)
data_q1_3_measures$BPXOSY <- rowMeans(data_q1_3_measures[, c("BPXOSY1", "BPXOSY2", "BPXOSY3")], na.rm = FALSE)
data_q1_3_measures$BPXODI <- rowMeans(data_q1_3_measures[, c("BPXODI1", "BPXODI2", "BPXODI3")], na.rm = FALSE)

data_q1_3_measures <- data_q1_3_measures %>%
  dplyr::select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, DR1TSODI, BPXOSY, BPXODI, DR1TALCO)
# Renaming columns
colnames(data_q1_3_measures) <- c("Sequence_no", "Gender", "Age", "Race", "Sodium_intake", "Systolic_BP", "Diastolic_BP", "Alcohol_consumption")

print(colSums(is.na(data_q1_3_measures)))
```

```{r, results='hide'}
data_q1 <- data.frame(data_q1_3_measures)


# 1. Impute Sodium_intake
data_for_model <- data_q1_3_measures %>%
  filter(!is.na(Sodium_intake), !is.na(Systolic_BP), !is.na(Diastolic_BP))

# Build the linear model
Sodium_intake_impute_model <- lm(Sodium_intake ~ Systolic_BP + Diastolic_BP + Age, data = data_for_model)

# Identify rows where Sodium_intake is missing
missing_Sodium_intake <- is.na(data_q1_3_measures$Sodium_intake)

# Filter rows with missing Sodium_intake but no missing predictors for prediction
rows_for_prediction <- data_q1_3_measures[missing_Sodium_intake, ] %>%
  filter(!is.na(Systolic_BP), !is.na(Diastolic_BP))

# Predict Sodium_intake for the rows where it can be predicted
predicted_Sodium_intake <- predict(Sodium_intake_impute_model, newdata = rows_for_prediction)

# Impute the predicted values back into the dataset
data_q1$Sodium_intake[missing_Sodium_intake] <- predicted_Sodium_intake



# 2. Impute Systolic_BP
data_for_model <- data_q1 %>%
  filter(!is.na(Sodium_intake), !is.na(Systolic_BP), !is.na(Alcohol_consumption))

# Build the linear model
Systolic_BP_impute_model <- lm(Systolic_BP ~ Sodium_intake + Alcohol_consumption + Age, data = data_for_model)

# Identify rows where Systolic_BP is missing
missing_Systolic_BP <- is.na(data_q1$Systolic_BP)

# Filter rows with missing Systolic_BP but no missing predictors for prediction
rows_for_prediction <- data_q1[missing_Systolic_BP, ] %>%
  filter(!is.na(Sodium_intake), !is.na(Alcohol_consumption))

# Predict Systolic_BP for the rows where it can be predicted
predicted_Systolic_BP <- predict(Systolic_BP_impute_model, newdata = rows_for_prediction)

# Impute the predicted values back into the dataset
data_q1$Systolic_BP[missing_Systolic_BP] <- predicted_Systolic_BP


# 3. Impute Diastolic_BP
data_for_model <- data_q1 %>%
  filter(!is.na(Sodium_intake), !is.na(Diastolic_BP), !is.na(Alcohol_consumption))

# Build the linear model
Diastolic_BP_impute_model <- lm(Diastolic_BP ~ Sodium_intake + Alcohol_consumption + Age, data = data_for_model)

# Identify rows where Diastolic_BP is missing
missing_Diastolic_BP <- is.na(data_q1$Diastolic_BP)

# Filter rows with missing Diastolic_BP but no missing predictors for prediction
rows_for_prediction <- data_q1[missing_Diastolic_BP, ] %>%
  filter(!is.na(Sodium_intake), !is.na(Alcohol_consumption))

# Predict Diastolic_BP for the rows where it can be predicted
predicted_Diastolic_BP <- predict(Diastolic_BP_impute_model, newdata = rows_for_prediction)

# Impute the predicted values back into the dataset
data_q1$Diastolic_BP[missing_Diastolic_BP] <- predicted_Diastolic_BP


# 4. Impute Alcohol_consumption
data_for_model <- data_q1_3_measures %>%
  filter(!is.na(Alcohol_consumption), !is.na(Systolic_BP), !is.na(Diastolic_BP))

# Build the linear model
Alcohol_consumption_impute_model <- lm(Alcohol_consumption ~ Systolic_BP + Diastolic_BP + Age, data = data_for_model)

# Identify rows where Alcohol_consumption is missing
missing_Alcohol_consumption <- is.na(data_q1_3_measures$Alcohol_consumption)

# Filter rows with missing Alcohol_consumption but no missing predictors for prediction
rows_for_prediction <- data_q1_3_measures[missing_Alcohol_consumption, ] %>%
  filter(!is.na(Systolic_BP), !is.na(Diastolic_BP))

# Predict Alcohol_consumption for the rows where it can be predicted
predicted_Alcohol_consumption <- predict(Alcohol_consumption_impute_model, newdata = rows_for_prediction)

# Impute the predicted values back into the dataset
data_q1$Alcohol_consumption[missing_Alcohol_consumption] <- predicted_Alcohol_consumption

# Verify no missing values remain
colSums(is.na(data_q1))
```

```{r, results='hide'}
# data_q1$BP_category <- with(data_q1, 
#   ifelse(Systolic_BP < 120 & Diastolic_BP < 80, "Normal",
#   ifelse(Systolic_BP >= 120 & Systolic_BP <= 129 & Diastolic_BP < 80, "Elevated",
#   ifelse((Systolic_BP >= 130 & Systolic_BP <= 139) | (Diastolic_BP >= 80 & Diastolic_BP <= 89), "Stage 1 Hypertension",
#   ifelse(Systolic_BP >= 140 | Diastolic_BP >= 90, "Stage 2 Hypertension", "Not Classified"))))
# )

data_q1$BP_category <- with(data_q1, dplyr::case_when(
  Systolic_BP <= 50 & Diastolic_BP <= 33 ~ "Dangerously low",
  Systolic_BP <= 60 & Diastolic_BP <= 40 ~ "Very low",
  Systolic_BP < 90 & Diastolic_BP < 60 ~ "Low",
  Systolic_BP < 120 & Diastolic_BP < 80 ~ "Normal",
  Systolic_BP >= 120 & Systolic_BP <= 129 & Diastolic_BP >= 80 ~ "Elevated",
  Systolic_BP >= 130 & Systolic_BP <= 139 & Diastolic_BP >= 80 & Diastolic_BP <= 89 ~ "Hypertension stage I",
  Systolic_BP >= 140 & Systolic_BP <= 159 & Diastolic_BP >= 90 & Diastolic_BP <= 99 ~ "Hypertension stage II",
  Systolic_BP >= 180 & Diastolic_BP >= 120 ~ "Hypertensive crisis",
  TRUE ~ "Not Classified"
))

# data_q1 <- data_q1 %>%
#   filter(data_q1$BP_category != "Not Classified")

#data_q1 <- data_q1[data_q1$BP_category != "Elevated", ]
```

### Model Fitting and Evaluation

To investigate the relationships between sodium intake and blood pressure categories, a multinomial logistic regression model was fitted. The initial model included main effects for sodium intake, age, gender, race, and alcohol consumption. Next, interaction terms between sodium intake and gender, as well as sodium intake and race, were introduced to determine whether the effect of sodium intake on blood pressure categories varied across these subgroups. This approach allowed us to explore potential effect modification by demographic factors.

Model comparisons were conducted using the Akaike Information Criterion (AIC), which evaluates the relative quality of statistical models by balancing goodness of fit and model complexity. Comparing the AIC values of the models with and without interaction terms helped determine whether the additional complexity introduced by the interactions was justified by an improved fit to the data.

A multiple linear regression (MLR) model was used for modeling blood mercury levels. Residual vs fitted plot was used to test linearity and homoscedasticity. R-squared was used to evaluate the performance of the linear regression model.

Variance Inflation Factors (VIFs) were calculated to detect multicollinearity among predictors. Variables with high VIFs (greater than 10) were considered for removal or transformation. Model fit was evaluated using AIC and deviance. A confusion matrix was used to evaluate the performance of the multinomial regression model. All statistical analyses and modeling were done using R programming language.

## Results

### Research Question 1: Contributing Factors of Blood Pressure Category

Exploratory data analysis (EDA) revealed several important insights. First, the sodium intake variable contained outliers with extremely high values, which could potentially skew the results. To address this, we limited sodium intake values to a maximum of 15,000 mg. This threshold was chosen based on the distribution of the data and practical dietary considerations.

Second, we observed that one of the racial groups had no observations in the dataset, leading to its exclusion from further analysis. Consequently, the analysis was restricted to six racial groups to ensure meaningful and interpretable results.

Third, we noted that the "Normal" blood pressure category had the highest number of observations among all categories. This imbalance suggests that the multinomial regression model may estimate higher probabilities for the "Normal" category compared to others, reflecting the underlying distribution of the data.

\[also add summary statistics here\]

```{r, results='hide'}
data_q1$Race <- as.factor(data_q1$Race)
data_q1$Race <- droplevels(data_q1$Race)

# Distribution of Sodium Intake (Boxplot)
ggplot(data_q1, aes(x = "", y = Sodium_intake)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  labs(title = "Distribution of Sodium Intake", x = "Sodium", y = "Sodium Intake (mg)") +
  theme_minimal()

# Distribution of Blood Pressure Categories (Bar chart)
ggplot(data_q1, aes(x = BP_category)) +
  geom_bar(fill = "lightgreen", color = "darkgreen") +
  labs(title = "Distribution of Blood Pressure Categories", x = "BP Category", y = "Count") +
  theme_minimal()

# # Distribution of Gender (Bar chart)
# ggplot(data_q1, aes(x = Gender)) +
#   geom_bar(fill = "lightpink", color = "darkred") +
#   labs(title = "Distribution of Gender", x = "Gender", y = "Count") +
#   theme_minimal()
# 
# # Distribution of Race (Bar chart)
# ggplot(data_q1, aes(x = Race)) +
#   geom_bar(fill = "lightblue", color = "darkblue") +
#   labs(title = "Distribution of Race", x = "Race", y = "Count") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   theme_minimal()

```

```{r, results='hide'}
continuous_vars <- data_q1 %>% 
  dplyr::select(Sodium_intake, Diastolic_BP, Systolic_BP, Alcohol_consumption, Age)

# Summary statistics for continuous variables (including min and max)
cat("Summary Statistics for Continuous Variables:\n")
continuous_summary <- data_q1 %>%
  summarise(
    across(
      c(Sodium_intake, Diastolic_BP, Systolic_BP, Alcohol_consumption, Age),
      list(min = min, max = max, mean = mean, sd = sd),
      na.rm = TRUE
    )
  )

print(continuous_summary)

# Detailed descriptive statistics using ⁠ skimr ⁠
cat("\nDetailed Descriptive Statistics:\n")
skim(continuous_vars)

# Summary statistics for categorical variables
categorical_vars <- data_q1 %>%
  dplyr::select(BP_category, Race, Gender)

cat("\nFrequency Tables for Categorical Variables:\n")
categorical_vars %>%
  summarise(across(everything(), ~ list(table(.)))) %>%
  t() %>%
  apply(1, function(x) {
    cat("\n", names(x), ":\n")
    print(as.data.frame(x[[1]]))
  })
```

| Variable Name            | Minimum | Maximum |    Mean | Standard Deviation |
|--------------------------|--------:|--------:|--------:|-------------------:|
| Sodium Intake            |       0 |   20006 | 3061.72 |            1411.03 |
| Diastolic Blood Pressure |      34 |     139 |   72.21 |              11.24 |
| Systolic Blood Pressure  |   71.33 |  232.33 |  119.04 |              17.86 |
| Alcohol Consumption      |   -0.66 |   448.1 |    6.02 |              17.68 |
| Age                      |       8 |      80 |   44.83 |              22.72 |

```{r, results='hide'}
# Remove outliers
data_q1 <- data_q1 %>% filter(data_q1$Sodium_intake < 15000)
max(data_q1$Sodium_intake)

# Ensure BP_category is an ordered factor
data_q1$BP_category <- factor(data_q1$BP_category, 
                         levels = c("Normal", "Elevated", "Stage 1 Hypertension", "Stage 2 Hypertension"), 
                         ordered = TRUE)
```

```{r, results='hide'}
# Convert categorical variables to factors
data_q1$Race <- factor(data_q1$Race, 
                  levels = c(1, 2, 3, 4, 6, 7),
                  labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White", 
                             "Non-Hispanic Black", "Non-Hispanic Asian", "Multiracial"))
```

```{r, results='hide'}
# Fit the multinomial logistic regression model without interaction terms
multinom_model <- multinom(factor(BP_category) ~ Sodium_intake + Age + Gender + Race + Alcohol_consumption, data = data_q1)

# Summary of the model
cat("\n********************Model Summary********************\n")
summary(multinom_model)

# Extracting p-values
z_values <- summary(multinom_model)$coefficients / summary(multinom_model)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))

cat("\n********************P-Values********************\n")
print(p_values)

# Predict probabilities for the dataset
predicted_probs <- predict(multinom_model, newdata = data_q1, type = "probs")

# Predict the category
predicted_categories <- predict(multinom_model, newdata = data_q1)

# Evaluate the model
# Confusion Matrix
cat("\n********************Confustion Matrix********************\n")
table(Predicted = predicted_categories, Actual = data_q1$BP_category)

# Print model diagnostics
cat("\n********************Model Coefficients:********************\n")
print(summary(multinom_model)$coefficients)

cat("\n********************Predicted Categories:********************\n")
head(predicted_categories)

cat("\n********************Predicted Probabilities:********************\n")
head(predicted_probs)

# # Calculate Variance Inflation Factors (VIF)
# vif_values <- vif(multinom_model)
# 
# # Display VIF values
# vif_values

cat("\n********************VIF Values********************\n")
vif_multinom <- function(multinom_model) {
  # Extract the design matrix (X) from the multinomial model
  X <- model.matrix(multinom_model)
  
  # Initialize a vector to store VIF values
  vif_values <- numeric(ncol(X) - 1)  # Exclude the intercept
  
  # Calculate VIF for each predictor
  for (i in 2:ncol(X)) {  # Start from 2 to skip the intercept
    # Regress the predictor against all others
    lm_model <- lm(X[, i] ~ X[, -i])
    
    # Compute VIF as 1 / (1 - R-squared)
    vif_values[i - 1] <- 1 / (1 - summary(lm_model)$r.squared)
  }
  
  # Assign variable names to the VIF values
  names(vif_values) <- colnames(X)[-1]
  
  return(vif_values)
}

# Run the VIF calculation on your multinomial model
vif_values <- vif_multinom(multinom_model)
print(vif_values)
```

```{r, results='hide'}
# Fit the multinomial logistic regression model with interaction terms
multinom_model <- multinom(BP_category ~ Sodium_intake*Gender + Sodium_intake*Race + Age + Gender + Race + Alcohol_consumption, data = data_q1)

# Summary of the model
cat("\n********************Model Summary********************\n")
summary(multinom_model)

# Extracting p-values
z_values <- summary(multinom_model)$coefficients / summary(multinom_model)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))

cat("\n********************P-Values********************\n")
print(p_values)

# Predict probabilities for the dataset
predicted_probs <- predict(multinom_model, newdata = data_q1, type = "probs")

# Predict the category
predicted_categories <- predict(multinom_model, newdata = data_q1)

# Evaluate the model
# Confusion Matrix
cat("\n********************Confusion Matrix********************\n")
table(Predicted = predicted_categories, Actual = data_q1$BP_category)

# Print model diagnostics
cat("\n********************Model Coefficients:********************\n")
print(summary(multinom_model)$coefficients)

cat("\n********************Predicted Categories:********************\n")
head(predicted_categories)

cat("\n********************Predicted Probabilities:********************\n")
head(predicted_probs)

# # Calculate Variance Inflation Factors (VIF)
# vif_values <- vif(multinom_model)
# 
# # Display VIF values
# vif_values

cat("\n********************VIF Values**************************\n")
vif_multinom <- function(multinom_model) {
  # Extract the design matrix (X) from the multinomial model
  X <- model.matrix(multinom_model)
  
  # Initialize a vector to store VIF values
  vif_values <- numeric(ncol(X) - 1)  # Exclude the intercept
  
  # Calculate VIF for each predictor
  for (i in 2:ncol(X)) {  # Start from 2 to skip the intercept
    # Regress the predictor against all others
    lm_model <- lm(X[, i] ~ X[, -i])
    
    # Compute VIF as 1 / (1 - R-squared)
    vif_values[i - 1] <- 1 / (1 - summary(lm_model)$r.squared)
  }
  
  # Assign variable names to the VIF values
  names(vif_values) <- colnames(X)[-1]
  
  return(vif_values)
}

# Run the VIF calculation on your multinomial model
vif_values <- vif_multinom(multinom_model)
print(vif_values)
```

The model's AIC of 12037.52 reflects a balance between complexity and fit. The confusion matrix shows the highest accuracy in predicting the "Normal" category but some misclassifications among hypertensive stages. The multinomial regression model revealed key associations between blood pressure categories and predictor variables. For "Elevated," the coefficient for sodium intake is negative (-0.000039), suggesting that increased sodium intake may be associated with reduced odds of being in this category compared to "Normal," though the effect size is very small. For "Stage 1" and "Stage 2 Hypertension," the coefficients are positive but small, indicating a slight increase in the odds of being in these categories with increased sodium intake. Age and alcohol consumption were positively associated with higher blood pressure categories, while gender and race demonstrated significant variation; for instance, females had lower odds of elevated and hypertensive categories, and "Non-Hispanic Black" was associated with higher odds of "Stage 1" and "Stage 2 Hypertension." Interaction terms highlighted nuanced relationships, such as weaker sodium intake effects for females but varying effects across racial groups. Predicted probabilities provided insights into classification confidence, with age and alcohol consistently influencing outcomes. While the model converged successfully, the modest effect sizes for some predictors and potential for further refinement suggest opportunities for further investigation.

### Research Question 2: Contributing Factors of Blood Mercury Level

The primary research question investigates the relationship between drinking water sources and blood mercury levels, while considering demographic factors. Specifically, we examined how tap water and bottled water consumption patterns influence blood mercury concentrations, and how these relationships might vary across different demographic groups. This investigation is particularly relevant given the growing public health concerns about environmental toxin exposure through different water sources.

Our statistical approach employed multiple linear regression with log-transformed mercury levels as the dependent variable, addressing both the non-normal distribution of mercury concentrations and the presence of detection limits in the measurements. The model incorporated water consumption variables (tap and bottled water intake), demographic characteristics (race, gender, age), and socioeconomic status (income-to-poverty ratio), along with key interaction terms to capture the complex relationships between these factors.

```{r data-2, warning=FALSE}

# Load datasets
dr1tot <- dietary
bpxo   <- BP
demo   <- demographics
pbcd   <- mercury

# Merge datasets on sequence number (SEQN)
data_merged <- demo %>%
  inner_join(bpxo, by = "SEQN") %>%
  inner_join(dr1tot, by = "SEQN") %>%
  inner_join(pbcd, by = "SEQN")

data <- dplyr::select(
  data_merged,
  SEQN, BPXOSY1, BPXODI1, RIDRETH3, RIAGENDR, RIDAGEYR, INDFMPIR, DR1_330Z, DR1_320Z, DR1BWATZ, 
  LBXTHG, LBDBPBSI, LBXBSE, DR1TSODI, DRD340, 
  starts_with("DRD350"), 
  starts_with("DRD370")
)


# Filter data for valid non-missing values

data$BloodPressureCategory <- cut(data$BPXOSY1, breaks = 
                                c(0, 50, 60, 90, 120, 129, 139, 140, 180, Inf),
      labels = c("Dange. Low", "Very Low", "Low", "Normal", "Elevated", 
             "HyperT Stg 1",  "HyperT Stg 2", "HyperT Crisis", "HyperT Crisis"))

# Create Gender variable
data$Gender <- factor(data$RIAGENDR, labels = c("Male", "Female"))

# Add a hypertension status column
data$Hypertensive <- ifelse(data$BPXOSY1 > 140 | data$BPXODI1 > 90, "Yes", "No")
data$Hypertensive <- as.factor(data$Hypertensive)

```

#### EDA

```{r plots-2b, echo=FALSE}
# Mecury levels
# Filter relevant columns
filtered_data <- data %>%
  dplyr::select(SEQN, DR1_330Z, DR1_320Z, DR1BWATZ, LBXTHG) %>%
  filter(!is.na(DR1_330Z), !is.na(DR1_320Z), !is.na(DR1BWATZ), !is.na(LBXTHG),
         LBXTHG < 40, LBXTHG > 6e-79)

# Reshape data
long_data <- filtered_data %>%
  pivot_longer(
    cols = c(DR1_330Z, DR1_320Z, DR1BWATZ),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  mutate(
    Variable = factor(Variable, levels = c("DR1_330Z", "DR1BWATZ", "DR1_320Z"), 
              labels = c("Tap Water", "Bottled Water", "Combined Plain Water"))
  )

# Create faceted scatter plot
plot2b1 <- ggplot(long_data, aes(x = Value, y = LBXTHG)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Effect of kinds of water consumed on Mercury Levels",
    x = "Water Intake (gm)",
    y = "Mercury Level (µg/L)"
  ) +
  theme_minimal()

plot2b2 <- ggplot(filtered_data, aes(x = LBXTHG)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Mercury Levels",
    x = "Mercury Level (µg/L)",
    y = "Frequency"
  ) +
  theme_minimal()

print(plot2b1)
# print(plot2b2)
```

1.  **Mercury Levels vs. Water Consumption** The faceted scatter plot examined the relationships between mercury levels and water intake across three types of water consumption (tap water, bottled water, and combined plain water). Linear trend lines showed that higher consumption of tap water might led to a slighly increase in blood mercury levels in people versus bottled water which had a no change in blood mecury levels as more is consumed.

```{r plots-3b, echo=FALSE}
# Filter relevant columns, including the updated race variable RIDRETH3
filtered_data <- data %>%
  dplyr::select(SEQN, DR1_330Z, DR1_320Z, DR1BWATZ, LBXTHG, RIDRETH3) %>%
  filter(!is.na(DR1_330Z), !is.na(DR1_320Z), !is.na(DR1BWATZ), 
         !is.na(LBXTHG), !is.na(RIDRETH3),
         LBXTHG < 40, LBXTHG > 6e-79)

# Map `RIDRETH3` to descriptive race labels
filtered_data <- filtered_data %>%
  mutate(
    Race = factor(RIDRETH3, levels = c(1, 2, 3, 4, 6, 7),
                  labels = c("Mexican American", "Other Hispanic", 
                             "Non-Hispanic White", "Non-Hispanic Black", 
                             "Non-Hispanic Asian", "Other Race - Including Multi-Racial"))
  )

# Create a boxplot to show mercury levels by race
plot_race <- ggplot(filtered_data, aes(x = Race, y = LBXTHG)) +
  geom_boxplot(fill = "skyblue", color = "black", outlier.color = "red", outlier.shape = 16) +
  labs(
    title = "Mercury Levels by Race/Ethnicity (Using RIDRETH3)",
    x = "Race/Ethnicity",
    y = "Mercury Level (µg/L)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create a violin plot for further visualization
plot_violin <- ggplot(filtered_data, aes(x = Race, y = LBXTHG, fill = Race)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_jitter(shape = 16, color = "black", size = 0.5, alpha = 0.5) +
  labs(
    title = "Distribution of Mercury Levels by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Mercury Level (µg/L)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Print the plots
#print(plot_race)
print(plot_violin)


```

2.  **Mercury Levels by Race/Ethnicity** The violin plot was used to compare mercury distributions across racial/ethnic groups. This graph shows that mercury levels vary significantly across racial and ethnic groups, with Non-Hispanic Asians having the highest levels and the widest distribution. We can see that most distribution are right-skewed meaning that most individuals in all racial cateogries have low mercury levels, but there are noticeable outliers, especially among Non-Hispanic Asians and Non-Hispanic Blacks.

```{r,include=FALSE}
# Pick only the columns of interest
data <- data_merged %>%
  dplyr::select(SEQN, RIAGENDR, RIDAGEYR, DR1_330Z, DR1_320Z, DR1BWATZ, 
         LBXTHG, RIDRETH3, INDFMPIR)

# data <- data %>%
#   filter(!is.na(SEQN),  !is.na(RIAGENDR), !is.na(RIDAGEYR),  !is.na(DR1_330Z), 
#          !is.na(DR1_320Z), !is.na(DR1BWATZ), !is.na(LBXTHG),
#          !is.na(RIDRETH3),!is.na(INDFMPIR))

data <- data %>%
  mutate(
    Race = factor(RIDRETH3, levels = c(1, 2, 3, 4, 6, 7),
                 labels = c("Mexican American", "Other Hispanic", 
                          "Non-Hispanic White", "Non-Hispanic Black", 
                          "Non-Hispanic Asian", "Other Race - Including Multi-Racial")),
    Gender = factor(RIAGENDR, levels = c(1, 2),
                   labels = c("Male", "Female")),
    Water_Type = case_when(
      DR1_330Z > 0 & DR1BWATZ > 0 ~ "Both",
      DR1_330Z > 0 & DR1BWATZ == 0 ~ "Tap Water",
      DR1_330Z == 0 & DR1BWATZ > 0 ~ "Bottled Water",
      TRUE ~ "None"
    ),
    Water_Type = factor(Water_Type, levels = c("Tap Water", "Bottled Water", "Both", "None"))
  )
```

```{r,include=FALSE}
# Stochastic regression imputations for missing values:
# 1. Impute tap water (DR1_330Z)
tap_water_model <- lm(DR1_330Z ~ Race + Gender + RIDAGEYR, 
                     data = subset(data, !is.na(DR1_330Z)))
missing_tap <- is.na(data$DR1_330Z)
predicted_tap <- predict(tap_water_model, newdata = data[missing_tap,])
set.seed(123)
tap_noise <- rnorm(n = length(predicted_tap), 
                  mean = 0, 
                  sd = summary(tap_water_model)$sigma)
data$DR1_330Z[missing_tap] <- predicted_tap + tap_noise

# 2. Impute bottled water (DR1BWATZ)
bottled_water_model <- lm(DR1BWATZ ~ DR1_330Z + Race + Gender + RIDAGEYR, 
                         data = subset(data, !is.na(DR1BWATZ)))
missing_bottled <- is.na(data$DR1BWATZ)
predicted_bottled <- predict(bottled_water_model, newdata = data[missing_bottled,])
bottled_noise <- rnorm(n = length(predicted_bottled), 
                      mean = 0, 
                      sd = summary(bottled_water_model)$sigma)
data$DR1BWATZ[missing_bottled] <- predicted_bottled + bottled_noise

# 3. Impute INDFMPIR
income_model <- lm(INDFMPIR ~ DR1_330Z + DR1BWATZ + Race + Gender + RIDAGEYR, 
                  data = subset(data, !is.na(INDFMPIR)))
missing_income <- is.na(data$INDFMPIR)
predicted_income <- predict(income_model, newdata = data[missing_income,])
income_noise <- rnorm(n = length(predicted_income), 
                     mean = 0, 
                     sd = summary(income_model)$sigma)
data$INDFMPIR[missing_income] <- predicted_income + income_noise

# 4. Impute LBXTHG
mercury_model <- lm(LBXTHG ~ DR1_330Z + DR1BWATZ + Race + Gender + RIDAGEYR + INDFMPIR, 
                   data = subset(data, !is.na(LBXTHG)))
missing_mercury <- is.na(data$LBXTHG)
predicted_mercury <- predict(mercury_model, newdata = data[missing_mercury,])
mercury_noise <- rnorm(n = length(predicted_mercury), 
                      mean = 0, 
                      sd = summary(mercury_model)$sigma)
data$LBXTHG[missing_mercury] <- predicted_mercury + mercury_noise

# Make sure all imputed values for LBXTHG are positive (since we'll take the log)
data$LBXTHG[data$LBXTHG <= 0] <- min(data$LBXTHG[data$LBXTHG > 0], na.rm = TRUE)

# Calculate log_LBXTHG after imputation
data$log_LBXTHG <- log(data$LBXTHG)

# Verify no missing values remain
colSums(is.na(data))
```

```{r,include=FALSE}
data$log_LBXTHG <- log(data$LBXTHG)

model_updated <- lm(log_LBXTHG ~ DR1_330Z + DR1BWATZ + Race*INDFMPIR + 
                   Gender + RIDAGEYR, 
                   data = data)

coef_table <- summary(model_updated)$coefficients
ci <- confint(model_updated)

characteristic <- c("Intercept",
                   "Tap Water Intake",
                   "Bottled Water Intake",
                   "Race: Other Hispanic",
                   "Race: Non-Hispanic White",
                   "Race: Non-Hispanic Black",
                   "Race: Non-Hispanic Asian",
                   "Race: Other/Multi-Racial",
                   "Income-to-Poverty Ratio",
                   "Gender: Female",
                   "Age (years)",
                   "Other Hispanic × Income",
                   "Non-Hispanic White × Income",
                   "Non-Hispanic Black × Income",
                   "Non-Hispanic Asian × Income",
                   "Other/Multi-Racial × Income")

beta <- sprintf("%.3f", coef_table[,1])
se <- sprintf("%.3f", coef_table[,2])
ci_values <- sprintf("%.3f, %.3f", ci[,1], ci[,2])
p_values <- ifelse(coef_table[,4] < 0.001, "< 0.001", sprintf("%.3f", coef_table[,4]))

# Combine into a matrix
results_matrix <- cbind(characteristic, beta, se, ci_values, p_values)
colnames(results_matrix) <- c("Predictor Variables", "Beta", "SE", "95% CI", "P-value")

# Create table
kable(results_matrix,
      align = c('l', 'c', 'c', 'c', 'c'),
      row.names = FALSE)

```

```{r, include=FALSE}
summary(model_updated)
```

The multiple linear regression model examines factors affecting blood mercury levels. Since the blood mercury level has a right-skewed distribution with a long right tail, we log-transformed the dependent variable for better model fit. The model is statistically significant (F-statistic = 81.04, p \< 0.001) and explains 13.51% of the variance in blood mercury levels with an R-squared of 0.1351. The model's low R-squared value is primarily because of two major reasons: (1) There are outstanding outliers of blood mercury levels, but our model's predictor variables on demographics and water drinking types are unlikely to be the key reasons that cause these extremely high mercury levels. The key reason that could lead to extremely high levels could be environmental factors, such as pollution. (2) For the mercury exam results, there is lower detection limit. When the actual mercury level is less than this detection limit of 0.21 µg/L, the variable would be recorded as 0.12, which is actual detection limit divided by the square root of 2. This creates a large cluster of identical values and true values below detection limit are unknown, which makes it harder for the model to explain variation that's artificially removed.

The model summary output shows that tap water intakes has a significant positive association with p-value less than 0.001, whereas bottled water intake has no significant effect (with p-value of 0.133). For the variable of race, compared to the reference level of Mexican Americans, non-Hispanic Asians show substantially higher mercury levels, non-Hispanic Whites show lower levels, other Hispanics show higher levels, and non-Hispanic Blacks and other races show no significant differences. For demographic factors, both age and income level have positive association with mercury levels (with p-value \< 0.001). Holding all other variables constant, females have lower mercury levels on average than males. For interaction term between race and income level, the positive interaction means the effect of income on mercury levels is stronger for non-Hispanic Whites compared to Mexican Americans. The negative interaction indicates that the effect of income on mercury levels is weaker for non-Hispanic Asians compared to Mexican Americans.

We also evaluated the residual plots of the model. For the residual plot, there is a straight line in the middle, which is caused by the detection limit of mercury level. For the Q-Q plot, the points follow the diagonal line closely, suggesting good normality in the central part of the distribution. However, there are some deviations in both tails, which are likely due to the outliers in the mercury level.

```{r, include=FALSE}

plot(model_updated)

```

## Conclusion

This report sheds light on some of the key factors influencing blood pressure and mercury levels in the U.S. population. While we didn’t find a significant link between dietary sodium intake and blood pressure, age stood out as a major predictor of hypertension. This suggests that efforts to prevent high blood pressure should focus more on age-related risk factors. When it comes to mercury exposure, tap water consumption, race, and income emerged as important contributors. Non-Hispanic Asian populations showed significantly higher mercury levels, which raises questions about whether dietary factors cause this or if there is influence from other relationships. To test this we created a model to determine the strength of factors on mercury levels in individuals. The complex relationship between race and income also suggests that environmental exposure risks are not evenly distributed, pointing to areas where public health policies could make a big difference. Overall, these findings suggest a more tailored approach to public health—one that considers individual and community-level differences in exposure and risk. By combining insights from diet, environment, and demographics, we can create more effective strategies to reduce hypertension and mercury toxicity in vulnerable populations.

This study does have some limitations, like the possibility of unmeasured factors and the reliance on observational data, which makes it harder to draw clear cause-and-effect conclusions. Future research could benefit from using long-term data and more gradular studies to better understand how individual and enviornmental factors drive disparities in mercury exposure and hypertension.

## References

Centers for Disease Control and Prevention. (2023). High blood pressure facts and statistics. <https://www.cdc.gov/highbloodpressure/data-research/facts-stats/index.html>

Houston, Michael C. "The Role of Mercury in Cardiovascular Disease." Journal of Clinical Hypertension, vol. 13, no. 8, 2011, pp. 621–627, https://doi.org/10.1111/j.1751-7176.2011.00489.x.

World Health Organization. "Hypertension." WHO, 2023, https://www.who.int/news-room/fact-sheets/detail/hypertension.

World Health Organization. (2021). Sodium intake and cardiovascular health. <https://www.who.int/news-room/fact-sheets/detail/sodium-intake-for-adults-and-children>

Agency for Toxic Substances and Disease Registry. (2023). Mercury and health. <https://www.atsdr.cdc.gov/toxprofiles/tp46.pdf>

## Appendix
